%% ========================================================================
%% EXAMPLE PRESENTATION WITH SLIDE REFERENCES
%% ========================================================================
%% This file demonstrates how to use the slide reference commands
%% created for formatting references in presentations.
%% ========================================================================

\documentclass[aspectratio=169]{beamer}

% Basic packages
\usepackage[utf8]{inputenc}
\usepackage[T1]{fontenc}
\usepackage{xparse}  % For enhanced command definitions (optional)

% Include the slide references commands
\input{slide_references_command.tex}

% Theme (choose your preferred one)
\usetheme{Madrid}
\usecolortheme{default}

% Title information
\title{Dynamic Graph Transformers for CVRP}
\subtitle{Reinforcement Learning Approach}
\author{Your Name}
\institute{Your Institution}
\date{\today}

\begin{document}

% Title slide (Slide 1) - no references
\begin{frame}
    \titlepage
\end{frame}

% Slide 2 - Basic usage with your example text
\begin{frame}
    \frametitle{Problem Formulation}
    
    \begin{itemize}
        \item The Capacitated Vehicle Routing Problem (CVRP) is a classical optimization challenge
        \item We propose a novel Graph Transformer approach
        \item Our method leverages reinforcement learning for dynamic route construction
    \end{itemize}
    
    \bigskip
    
    Graph neural networks have shown promising results in combinatorial optimization tasks.
    
    % Your example references with basic command
    \sliderefs{Weinberg (1994), Heggie+ (2020), Landau (1946), van Kampen (1955), Case (1959), Ng et al. (1999, 2004); Ng \& Bhattacharjee (2021)}
    
\end{frame}

% Slide 3 - Using smaller text version
\begin{frame}
    \frametitle{Graph Transformer Architecture}
    
    \begin{columns}
        \begin{column}{0.6\textwidth}
            \begin{itemize}
                \item Multi-head attention mechanism
                \item Node and edge embeddings
                \item Dynamic graph construction
            \end{itemize}
        \end{column}
        \begin{column}{0.4\textwidth}
            % Add your figure here
            \begin{center}
                [Architecture Diagram]
            \end{center}
        \end{column}
    \end{columns}
    
    % Using even smaller text version
    \sliderefsSmaller{Transformer architectures have revolutionized sequence modeling tasks. Recent work extends these concepts to graph-structured data with impressive results.}
    
\end{frame}

% Slide 4 - Using custom line length
\begin{frame}
    \frametitle{Reinforcement Learning Framework}
    
    Our approach treats route construction as a sequential decision-making process:
    
    \begin{enumerate}
        \item State representation using graph embeddings
        \item Action space defined by customer selection
        \item Reward based on tour improvement
    \end{enumerate}
    
    The policy network learns to make optimal routing decisions through trial and error.
    
    % Using custom line length (4cm instead of default 3.5cm)
    \sliderefsCustom[4cm]{Policy gradient methods provide stable learning for discrete action spaces in combinatorial problems.}
    
\end{frame}

% Slide 5 - Bottom-aligned references
\begin{frame}
    \frametitle{Experimental Results}
    
    \begin{itemize}
        \item Tested on standard CVRP benchmarks
        \item Compared against state-of-the-art methods
        \item Achieved competitive performance
    \end{itemize}
    
    \begin{table}[h]
        \centering
        \small
        \begin{tabular}{|l|c|c|}
        \hline
        \textbf{Method} & \textbf{Cost} & \textbf{Time} \\
        \hline
        OR-Tools & 0.425 & 58s \\
        Our Method & \textbf{0.420} & \textbf{12s} \\
        \hline
        \end{tabular}
        \caption{Performance comparison on N=10 instances}
    \end{table}
    
    % Bottom-aligned references
    \sliderefsBottom{Benchmark results demonstrate significant improvements in both solution quality and computational efficiency.}
    
\end{frame}

% Slide 6 - Using longer line
\begin{frame}
    \frametitle{Computational Efficiency}
    
    \begin{columns}
        \begin{column}{0.5\textwidth}
            \textbf{GPU Implementation:}
            \begin{itemize}
                \item Parallel graph processing
                \item Batch policy evaluation
                \item Memory-efficient attention
            \end{itemize}
        \end{column}
        \begin{column}{0.5\textwidth}
            \textbf{Performance Gains:}
            \begin{itemize}
                \item 10x speedup vs CPU
                \item Scales to larger instances
                \item Real-time applications
            \end{itemize}
        \end{column}
    \end{columns}
    
    % Using longer line version
    \sliderefsLong{GPU acceleration enables handling of large-scale routing problems that were previously computationally prohibitive.}
    
\end{frame}

% Slide 7 - Regular references again
\begin{frame}
    \frametitle{Future Directions}
    
    \begin{itemize}
        \item Extension to multi-objective optimization
        \item Integration with real-world constraints
        \item Online learning for dynamic environments
    \end{itemize}
    
    \bigskip
    
    The framework provides a foundation for addressing more complex variants of vehicle routing problems.
    
    % Back to regular references
    \sliderefs{Dynamic vehicle routing presents significant challenges for logistics optimization in changing environments.}
    
\end{frame}

% Slide 8 - Conclusion
\begin{frame}
    \frametitle{Conclusion}
    
    \textbf{Key Contributions:}
    \begin{enumerate}
        \item Novel Graph Transformer architecture for CVRP
        \item Efficient GPU implementation
        \item Competitive results on standard benchmarks
    \end{enumerate}
    
    \bigskip
    
    \textbf{Thank you for your attention!}
    
    % Final slide references
    \sliderefs{This work opens new possibilities for applying deep learning to classical optimization problems.}
    
\end{frame}

\end{document}
